Efficient pedigree recording for fast population genetics simulation

### 2017-12-09 forwards_paper.tex
Peter edited through

\begin{abstract}
    The computational burden of tracking individual genomes
    in a population genetics simulation can be substantial.
    In this paper we describe how to dramatically reduce this burden by
    efficiently recording the entire history of the population.
    We do this by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording the entire history of genetic inheritance in an efficient representation of the ancestral recombination graph,
    on which neutral mutations can be quickly placed afterwards ---
    the \emph{tree sequence} introduced in the software package \msprime.
    We implement the method in two popular fowards-time simulation frameworks,
    and show it leads to order-of-magnitude speedups.
    % This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
\end{abstract}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g.\ $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of
(a) random mating,
(b) neutrality,
and (c) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent,
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical space
and/or have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, allee effects, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
The third assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
Several studies have now shown that in samples of size approaching that of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of
interest in large scale forwards-time, individual-based simulations.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
\cite{Sanjak2017-ko} used
fwdpp \citep{fwdpp} to simulate a series of models of quantiative traits under mutation-selection balance with
population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\approx 5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

Modern computing power easily allows simulations of birth, death and reproduction
in a population having even hundreds of millions of individuals,
even though theory tells us that a population of size $N$
must be run for many multiples of $N$ generations to produce stable genetic patterns.
However, if our interest lies in the resulting genetic patterns of variation
-- and often, the point of such simulations is to compare to real data --
then such simulations must somehow produce at the end data for each individual on a genomic scale.
As samples of most species' genomes harbor tens or hundreds of millions of variant sites,
naively carrying full genotypes for even modest numbers of individuals through a simulation
becomes quickly prohibitive.
Because of this computational burden of neutral mutations, even the fastest simulation frameworks such as SLiM,
SLiM2~\citep{haller2017flexible} and fwdpp~\citep{fwdpp}
can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
for tens of thousands of generations, and there are still near-order-of-magnitude performance differences between these
implementations~\citep{haller2017flexible}.
In practice, current state-of-the-art simulation software may take on the order of
weeks to simulate models without selection~\citep{fwdpp,Hernandez2015-wf},
and different simulation engines differ in how efficiently they
calculate fitnesses in models with selection~\citep{fwdpp}.
These population and region sizes are still substantially short of whole genomes
(hundreds to thousands of megabases)
for many biological population sizes of interest.

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, the alleles carried by individuals in a population at neutral sites
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
simply laying down neutral mutations on top of that history afterwards
is equivalent to having generated them during the simulation.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
In this paper, we describe how algorithmic tools and data structures developed for the
coalescent simulator \msprime{}
can be used to efficiently record, and later process, this history.
In so doing we record the \emph{population pedigree} --
the complete history of parent-offspring relationships of an entire population
going back to a remote time including the genetic outcomes of each ancestral
meiosis.
This embellished pedigree contains all the information necessary
to construct the genealogical tree that relates each individual to each other
at each position on the genome, i.e., the \emph{tree
sequence}~\citep{kelleher2016efficient}.
Combined with ancestral genotypes and the origins of new mutations,
it also completely specifies the genomic sequence of any individual in the population at any time.
This is much more than we need to know, however,
so we discard all information irrelevant to the genetic history
of the \emph{sampled} individuals,
which results in considerable savings.
Another way of representing this same information
is known as the \emph{ancestral recombination graph} ---
the {ARG} \citep{griffiths1997ancestral},
where nodes are common ancestor or recombination events ---
which has been the subject of substantial study
under the assumptions of coalescent
theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.

\jk{And finish with a more detailed descrition of what we do in the paper}
In this paper we discuss storage methods for genealogies (and hence, genome sequence),
an algorithm for simplifying these,
and their use in forwards-time simulation.
Although we were motivated by a need for more efficient simulations,
these tools may prove more widely useful.

@@

We begin by benchmarking the performance improvement achieved by connecting
two previously published forwards-time simulation libraries,
\simupop{} \citep{simupop} and \fwdpp{} \citep{fwdpp}
to the tools we describe here, implemented in \msprime{}.
Then, we describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences in several \emph{tables};
(b) an algorithm, \algref{W}, to record tree sequence information into these tables on the fly
    from a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
We then describe a general-purpose method that uses these tools
(as implemented in \msprime{})
to efficiently record simplified a tree sequence,
and analyze of its run time complexity.

@@

To measure the performance gains from ancestry recording, we ran simulations implemented in C++ (see below) with and
without ancestry tracking.  The simulations without ancestry tracking introduced neutral mutations at a scaled mutation
rate equal to the scaled recombination rate (\textit{e.g.}, $\theta = 4N\mu = \rho$, where $\rho = 4Nr$, $N$ is the
diploid population size, $\mu$ is the
mutation rate to neutral mutations (per gamete, per generation) and $r$ is the recombination rate per diploid per
generation.  Both $\mu$ and $r$ are the means of a Poisson process and we used the infinitely-many sites mutation model
\citep{Kimura1969-uc}.

We varied $N \in (1000, 10000, 50000)$, $\rho \in (10^3, 10^4, 10^5)$ and ran a single replicate of all $(N,\rho)$
combinations.  Deleterious mutations were introduced at rate $\rho/100$ per generation, drawing scaled selection
coefficients ($2Ns$)
from a Gamma distribution with a mean of -5 and a shape parameter of 1.  This distribution of fitness effects results in
thousands of weakly-deleterious mutations segregating in the population, many of which drift to intermediate
frequencies.  The case of many mutations with selection is a non-trivial violation of exchangeability assumptions of the
coalescent \citep{Neuhauser1997-nn}.  Therefore, these selected mutations must be explicitly tracked in our forward simulation
and the time savings due to ancestry tracking come from not having to record \textit{neutral} mutations.

We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 Gigahertz Intel E5-2650 CPU with
hyperthreading enabled.  We ran one simulation at a time and the machine was under minimal load otherwise.  We used GNU
parallel \citep{Tange2011a} to kill
any simulation that did not finish within 72 hours.  We used the Linux \texttt{time} command to record run time and peak
memory usage of each replicate and our simulation scripts separately recorded the time spent in various steps of each
run.

@@

@@

A \emph{tree sequence} is an encoding for a sequence of correlated trees,
such as those describing the history of a sexual population.
It is efficient because branches that are shared by adjacent trees are stored once,
rather than repeatedly for each tree.
The topology of a tree sequence is defined via its \emph{nodes} and \emph{edges},
while information about variants are recorded as \emph{sites} and \emph{mutations}.
The format for these four tables is given in more detail in the Methods;
here we explain conceptually,
using the example of Figure~\ref{fig:example_tree_sequence}. This formulation
is derived from the ``coalescence records'' encoding of tree
sequences~\citep{kelleher2016efficient}, normalised to remove redundancy
and generalised to include a much larger class of tree topology.

The \emph{nodes} of a tree sequence
correspond to the vertices in the individual genealogies along the sequence,
and each node refers to a distinct ancestor.
Since each node represents a specific ancestor, it has a unique ``time'',
thought of as her birth time, which determines the height of any vertices
she is associated with.
Note that since each node time is equal to the amount of time since the {birth} of the
corresponding parent, time is measured in clock time, not in meioses.
The example of Figure~\ref{fig:example_tree_sequence} has five nodes:
nodes 0, 1 and 2 occur at time 0 and are the \emph{samples},
while nodes 3 and 4 represent those ancestors necessary to record their genealogy,
who were born at time 1 and 2 respectively.

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10.
        The left-hand panel shows the tree sequence pictorially in two different ways:
        (top) as a sequence of tree topologies
        and (bottom) the spatial extent of the edges that define these topologies.
        The right-hand panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations)
        defined by \msprime.
        \plr{Whoops!  genotype of 1 in the second tree should be `G'!}
        \jk{Some minor alignment issues present here.}
        \plr{turns out this is no fun to edit in inkscape.
            I can re-make to match the other figs
            but will leave alone for now: it's good!}
        \label{fig:example_tree_sequence}
    }
\end{figure}

The \emph{edges} define how nodes relate to each other over specific genomic intervals.
Each edge records:
% is a tuple $(\ell, r, p, c)$, where
the endpoints $[\ell, r)$ of the half-open genomic interval defining the
spatial extent of the edge;
and the identities $p$ and $c$ of the parent and child nodes
of a single branch that occurs in all trees covering this interval.
The spatial extent of the edges defining the topology of Figure~\ref{fig:example_tree_sequence}
are shown in the bottom left panel.
For example, the branch joining nodes 1 to 3 appears both trees,
and so is recorded as a single edge extending over the whole chromosome.
It is this method of capturing the shared structure between adjacent trees that makes the
tree sequence encoding very compact and algorithmically efficient.

Recovering the sequence of trees from this information is straightforward:
each point along the genome that the tree topology changes
is accompanied by the end of some \emph{edges} and the beginning of others.
Since each \emph{edge} records the genomic interval
over which a given node inherits from a particular ancestor,
to construct the tree at a certain point in the genome
we need only retrieve all edges overlapping that point
and construct the corresponding tree.
To modify the tree to reflect the genealogy at a nearby location,
we simply remove those edges whose intervals do not overlap that location,
and add those new edges whose intervals do.
Incidentally, this property that edges naturally encode \emph{differences}
between nearby trees (e.g., as ``subtree prune and regraft'' moves)
allows for efficient algorithms that take advantage
of the highly correlated nature of nearby trees~\citep{kelleher2016efficient}.

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
is associated with a position on the genome and an ancestral state. For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2 with ancestral state `A' and the other at position 7 with ancestral state `G'. If
no mutations occur, all samples inherit the ancestral state at a given site.
Each mutation occurs above a specific node at a given site,
and results in a specific derived state.
Thus, all samples below the mutation node in the tree will inherit this state
(unless further mutations are encountered).
Three mutations are shown in Figure~\ref{fig:example_tree_sequence},
illustrated by red hexagons.
The first mutation occurs at site zero (in the left-hand tree), and is a simple
mutation resulting in node $2$ inheriting the state `T'.
The second site (in the right hand tree) has two mutations:
one occurring over node $3$ changing the state to `C',
and a back mutation over node $1$ changing the state to `G'.

@@

\paragraph{Validity of a set of tables}
Given a set of node and edge tables as described above,
there are a small set of requirements that ensure the tables
describe a valid tree sequence.
% (There are essentially no such requirements on the site and mutation tables.)
These are:
\begin{enumerate}
    \item Offspring must be born after their parents (and hence, no loops).
    \item The set of intervals on which each individual is a child must be disjoint.
\end{enumerate}
A pair of node and edge tables that satisfy these two requirements
is guarenteed to uniquely describe at each point on the genome
a collection of directed, acyclic graphs -- in other words, a forest of trees.
For some applications it is necessary to check that at every point
there is only a \emph{single} tree.
Checking this is more difficult, but is implemented in \msprime{}'s API.
\jda{TODO add function name}
For efficiency, \msprime{} makes several other sortedness requirements on the tables,
that are not necessarily satisfied by tables emitted by a forwards-time simulation.
\msprime{}'s API includes tools to rectify this by first sorting (using \texttt{sort\_tables})
and then using \texttt{simplify}, which works on sorted tables
and is guarenteed to produce a valid, \msprime{}-ready tree sequence.

@@

To record the genealogical history of a forwards in time simulation
we need to record two things for each new chromosome:
the birth time, and the endpoints and parental IDs of each distinctly inherited segment,
which are naturally stored as the \emph{nodes} and \emph{edges} of a tree sequence.
For concreteness, here we write out in pseudocode how to run a neutral Wright--Fisher simulation
(but with overlapping generations) that records genealogical history in this way.
The simulation will run for $T$ generations,
and has $N$ haploid individuals, each carrying a single chromosome of length L.
For simplicity we sample exactly one crossover per generation.
The probability of death per individual each generation is $\delta$.
We will use
$\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random
(and each such instance represents and independent draw). Given a node table
$\Nt$, the function $\taddnode{\Nt, t}$ adds a new node to the table with time $t$
and returns the ID of this new node. Similarly, the function $\taddedge{\Et,
\ell, r, p, c}$ adds a new edge $(\ell, r, p, c)$ to the edge table $\Et$.
The function $\tsimplify{P, \Nt, \Et}$ simplifies the history stored in the
tables $\Nt$ and $\Et$ to the minimal information required to represent the
genealogies of the list of node IDs $P$ (see section XXX).
We first list
the algorithm in detail, and immediately follow this with a step-by-step
explanation.

\begin{taocpalg}{W}{Forwards-time tree sequence}
{Simulates a randomly mating population of $N$ haploid individuals with
chromosome of length $L$ for $T$ generations, and returns the node
and edge tables ($\Nt$ and $\Et$) recording the simulated history.
Simplify is run every $s$ generations, removing genealogical
information from $\Nt$ and $\Et$ unreachable from the current population $P$.
}

\algstep{W1.}{Initialisation.}{
Set $\Nt \leftarrow \mbox{NodeTable}()$, $\mathcal{E}
\leftarrow \mbox{EdgeTable}()$ and $t \leftarrow T$.
 For $0 \leq j < N$, set $P_j \leftarrow \taddnode{\Nt, T}$. Set $j \leftarrow 0$.
}

\algstep{W2.}{New node.}{Set $u \leftarrow \taddnode{\Nt, t}$ and $P'_j \leftarrow u$.
}

@@

\algstep{W6.}{Simplify.}{Call $\tsimplify{P', \Nt, \Et}$, and set $P'_j
\leftarrow j $ for $0 \leq j < N$. (Tables may need to be sorted.)}

@@

We begin in~\algref{W1} by creating new node and edge tables, and setting
our population $P$ (a vector of $N$ node IDs) to the initial population.
This initial population is a set of $N$ nodes with birth time $T$ generations
ago. We also initialise our generation clock $t$ and individual index $j$.
Steps~\algref{W2}--\algref{W4} replace the individual with node ID
$P_j$ by first creating a new node with birth time $t$ (and ID $u$).
Then, in step~\algref{W3} we choose two indexes $a$ and $b$ uniformly (giving us
parental IDs $P_a$ and $P_b$) and choose a breakpoint $x$. We record the effects of this
event by storing two new edges: one recording that the parent of node $u$
from $0$ to $x$ is $P_a$, and another recording that the parent of $u$
from $x$ to $L$ is $P_b$. Step \algref{W5} then iterates over each of the
$N$ individuals for each generation. At the end of a generation, we then check
if we need to simplify (which is done every $s$ generations). If simplification
is required, we do this in step \algref{W6} by calling the simplify function
on the node and edge tables with the current set of population IDs $P'$ as
the samples. This updates the tables in-place to remove all redundant
information, and remaps the specified sample IDs to $0, \dots N - 1$
in the updated tables. Hence, we set our current population IDs to
$0, \dots N - 1$ after simplify has completed. (See section XXX for more
details on simplify.) Step \algref{W7} then completes the algorithm by looping
over generations; we decrement our clock $t$, and terminate if $t = 0$.
Otherwise, we update our current population and individual index and loop
back to \algref{W2}.

>>>>

\begin{abstract}
    In this paper we describe how to 
    efficiently record the entire genetic history of a population
    during a forwards-time, individual-based population genetics simulation.
    This allows us to dramatically reduce the computational burden of tracking individual genomes
    by simulating only those loci that may affect reproduction (those having non-neutral variants),
    and recording history in the \emph{tree sequence} format introduced in the software package \msprime,
    on which neutral mutations can be quickly placed afterwards.
    We implement the method in two popular forwards-time simulation frameworks,
    and show that not only does this speed up large simulations by one or two orders of magnitude,
    but saves substantially more information from the simulation
    (all marginal genealogies).
    % This has the promise of making large-scale, whole-genome simulations with realistic geography and selection finally possible.
\end{abstract}

@@

Since the 1980's, coalescent theory has enabled computer simulation of the results of population genetics models
identical to that which would be produced by large, randomly mating populations over long periods of time
without actually requiring simulation of so many generations or meioses.
Coalescent theory thus had three transformative effects on population genetics:
first, giving researchers better conceptual tools to describe \emph{gene trees} and thus bringing within-population trees into better focus;
second, producing analytical methods to estimate parameters of interest from genetic data (e.g., $\theta = 4N_e \mu$);
and finally, providing a computationally feasible method to produce computer simulations of population genetics processes.
However, these powerful advances came with substantial caveats:
the backwards-in-time processes that are described by coalescent theory
are only \emph{Markovian}, and thus feasible to work with,
thanks to the important assumptions of
(a) random mating,
(b) neutrality,
and (c) small sample size relative to the population size.
The first two assumptions can be side-stepped to a limited extent,
but it remains a challenge to map results of coalescent models
onto species that are distributed across continuous geographical space
and/or have large numbers of loci under various sorts of selection.
Furthermore, the relationship between the life history of a species --
fecundity and mortality schedules, Allee effects, and demographic fluctuations --
are all absorbed into a single compound parameter, the coalescence rate.
The third assumption is no longer safe, either --
for example, a recent study~\citep{martin2017human}
simulated 600,000 samples of human chromosome 20 to examine biases in GWAS.
Several studies have now shown that in samples of size approaching that of the population,
genealogical properties may be distorted relative to the coalescent expectation
\citep{wakeley2003gene,maruvka2011recovering,bhaskar2014distortion}.
These considerations, and increasing computational power, have led to a resurgence of
interest in large forwards-time, individual-based simulations.
For instance, \citet{harris2016genetic} used SLiM \citep{slim} to simulate ten thousand human exomes
to assess the impact of genetic load and Neanderthal introgression on human genetic diversity.
\cite{Sanjak2017-ko} used
fwdpp \citep{fwdpp} to simulate a series of models of quantitative traits under mutation-selection balance with
population sizes of $2 \times 10^4$ diploids in stable populations and populations growing up to $\approx 5
\times 10^5$ individuals, using the output to explore the relationship between the genotype/phenotype model and GWAS
outcomes.

Modern computing power easily allows simulations of birth, death and reproduction
in a population having even hundreds of millions of individuals.
However, if our interest lies in the resulting genetic patterns of variation
-- and often, the point of such simulations is to compare to real data --
then such simulations must somehow produce at the end data for each individual on a genomic scale.
As samples of most species' genomes harbor tens or hundreds of millions of variant sites,
naively carrying full genotypes for even modest numbers of individuals through a simulation
becomes quickly prohibitive.
To make matters worse, 
a population of size $N$ must be simulated across many multiples of $N$ generations
to produce stable genetic patterns.
Because of this computational burden of neutral mutations, even the fastest simulation frameworks such as SLiM,
SLiM2~\citep{haller2017flexible} and fwdpp~\citep{fwdpp}
can ``only'' simulate tens of megabases of sequence in tens of thousands of individuals
for tens of thousands of generations, and there are still near-order-of-magnitude performance differences between these
implementations~\citep{haller2017flexible}.
In practice, current state-of-the-art simulation software may take on the order of
weeks to simulate models without selection~\citep{fwdpp,Hernandez2015-wf},
and existing simulation engines differ in how efficiently they
calculate fitnesses in models with selection~\citep{fwdpp}.
These population and region sizes are still substantially short of whole genomes
(hundreds to thousands of megabases)
for many biological population sizes of interest.

However, it is thought that most genetic variation is selectively neutral (or nearly so).
By definition, the alleles carried by individuals in a population at neutral sites
do not affect the population process.
For this reason, if one records the entire genealogical history of a population over the course of a simulation,
simply laying down neutral mutations on top of that history afterwards
is equivalent to having generated them during the simulation.
Precisely, we would need to know the genealogical tree relating all sampled individuals
at each position along the genome.
In this paper, we describe how algorithmic tools and data structures developed for the
coalescent simulator \msprime{}
can be used to efficiently record, and later process, this history.
In so doing we record the \emph{population pedigree} --
the complete history of parent-offspring relationships of an entire population
going back to a remote time including the genetic outcomes of each ancestral
meiosis.
This embellished pedigree contains all the information necessary
to construct the genealogical tree that relates each individual to each other
at each position on the genome, i.e., the \emph{tree sequence}~\citep{kelleher2016efficient}.
Combined with ancestral genotypes and the origins of new mutations,
it also completely specifies the genomic sequence of any individual in the population at any time.
This is much more than we need to know, however,
so we discard all information irrelevant to the genetic history
of the \emph{sampled} individuals,
which results in considerable savings.
Another way of representing this same information
is known as the \emph{ancestral recombination graph} --
the {ARG} \citep{griffiths1997ancestral},
where nodes are common ancestor or recombination events --
which has been the subject of substantial study
under the assumptions of coalescent
theory~\citep{wiuf1997number,wiuf1999ancestry,marjoram2006coalescent,wilton2015smc}.
We use tree sequences for the substantial algorithmic advantages exploited in~\citep{kelleher2016efficient}.

In this paper we discuss storage methods for genealogies (and hence, genome sequence),
an algorithm for simplifying these,
and their use in forwards-time simulation.
Although we were motivated by a need for more efficient genomic simulations,
these tools may prove more widely useful.

@@

We begin by benchmarking the performance improvement achieved by connecting
two previously published forwards-time simulation libraries,
\simupop{} \citep{peng2005simupop} and \fwdpp{} \citep{fwdpp}
to the tools we describe here, as implemented in \msprime{}.
Then, we describe the conceptual and algorithmic foundations for the method:
(a) a format, implemented in the \msprime{} Python API,
for recording tree sequences in several \emph{tables};
(b) an algorithm to record tree sequence information into these tables on the fly
    from a forwards-time simulation;
and (c) an algorithm to \emph{simplify} a tree sequence, i.e., remove redundant information.
We then describe a general-purpose method that uses these tools
(as implemented in \msprime{})
to efficiently record simplified a tree sequence,
and analyze its run time complexity.

@@

To measure the performance gains from ancestry recording, we ran simulations implemented in C++ (see below) with and without ancestry tracking.  
The simulations without ancestry tracking introduced neutral mutations at a rate 
equal to the recombination rate,
so $\mu = r$, where $\mu$ and $r$ are the expected number of mutations and recombinations,
respectively, per gamete and per generation.
Both mutations and recombinations occur as Poisson processes with respective means $\mu$ and $r$,
and we used the infinitely-many sites mutation model \citep{Kimura1969-uc}.
We categorize simulations in terms of the \emph{scaled} parameters
$\theta = 4N\mu$ and $\rho = 4Nr$, where $N$ is the diploid population size,
since these most strongly determine run time.

We ran one simulation for each combination of $N$ from $(1000, 10000, 50000)$, and $\rho$ from $(10^3, 10^4, 10^5)$.
Deleterious mutations were introduced at rate $\rho/100$ per generation, 
drawing scaled selection coefficients ($2Ns$)
from a Gamma distribution with a mean of -5 and a shape parameter of 1.  
This distribution of fitness effects results in
thousands of weakly-deleterious mutations segregating in the population, 
many of which drift to intermediate frequencies.  
The case of many mutations with selection is a non-trivial violation of exchangeability assumptions of the
coalescent \citep{Neuhauser1997-nn}.
Therefore, these selected mutations must be explicitly tracked in our forward simulation,
and the time savings due to ancestry tracking come from not having to record \textit{neutral} mutations.

We ran all benchmarks on an Ubuntu Linux (version 16.04) system with two 2.6 Gigahertz Intel E5-2650 CPU with
hyperthreading enabled.  
We ran one simulation at a time and the machine was under minimal load otherwise.  
We used GNU parallel \citep{Tange2011a} to kill any simulation that did not finish within 72 hours,
and the Linux \texttt{time} command to record run time and peak memory usage of each replicate.
Our simulation scripts separately recorded the time spent in various steps of each run.

@@

\plr{Insert figures/tables of performance here and describe them.}

@@

The simulations above recorded information about each new individual 
in collection of tables that together encode a \emph{tree sequence}.
A {tree sequence} is an encoding for a sequence of correlated trees,
such as those describing the history of a sexual population.
It is efficient because branches that are shared by adjacent trees are stored once,
rather than repeatedly for each tree.
The topology of a tree sequence is defined via its \emph{nodes} and \emph{edges},
while information about variants are recorded as \emph{sites} and \emph{mutations}.
The format for these four tables is given in more detail in the Methods;
here we explain conceptually,
using the example of Figure~\ref{fig:example_tree_sequence}. This formulation
is derived from the ``coalescence records'' encoding of tree
sequences~\citep{kelleher2016efficient}, normalised to remove redundancy
and generalised to include a much larger class of tree topology.

The \emph{nodes} of a tree sequence
correspond to the vertices in the individual genealogies along the sequence,
and each node refers to a distinct ancestor.
Since each node represents a specific ancestor, it has a unique ``time'',
thought of as her birth time, which determines the height of any vertices
she is associated with.
Note that since each node time is equal to the amount of time since the {birth} of the
corresponding parent, time is measured in clock time, not in meioses.
The example of Figure~\ref{fig:example_tree_sequence} has five nodes:
nodes 0, 1 and 2 occur at time 0 and are the \emph{samples},
while nodes 3 and 4 represent those ancestors necessary to record their genealogy,
who were born one and two units of time in the past, respectively.

\begin{figure}
    \begin{center}
        \includegraphics[width=\textwidth]{example_tree_sequence}
    \end{center}
    \caption{
        An example tree sequence with three samples over a chromosome of length 10.
        The left-hand panel shows the tree sequence pictorially in two different ways:
        (top) as a sequence of tree topologies
        and (bottom) the spatial extent of the edges that define these topologies.
        The right-hand panels show the specific encoding
        of this tree sequence in the four tables (nodes, edges, sites and mutations)
        defined by \msprime.
        \plr{Whoops!  Genotypes in the second tree should be `GGC'}
        \jk{Some minor alignment issues present here.}
        \plr{turns out this is no fun to edit in inkscape.
            I can re-make to match the other figs
            but will leave alone for now: it's good!}
        \label{fig:example_tree_sequence}
    }
\end{figure}

The \emph{edges} define how nodes relate to each other over specific genomic intervals.
Each edge records:
% is a tuple $(\ell, r, p, c)$, where
the endpoints $[\ell, r)$ of the half-open genomic interval defining the
spatial extent of the edge;
and the identities $p$ and $c$ of the parent and child nodes
of a single branch that occurs in all trees covering this interval.
The spatial extent of the edges defining the topology of Figure~\ref{fig:example_tree_sequence}
are shown in the bottom left panel.
For example, the branch joining nodes 1 to 3 appears in both trees,
and so is recorded as a single edge extending over the whole chromosome.
It is this method of capturing the shared structure between adjacent trees that makes the
tree sequence encoding very compact and algorithmically efficient.

Recovering the sequence of trees from this information is straightforward:
each point along the genome that the tree topology changes
is accompanied by the end of some {edges} and the beginning of others.
Since each {edge} records the genomic interval
over which a given node inherits from a particular ancestor,
to construct the tree at a certain point in the genome
we need only retrieve all edges overlapping that point
and construct the corresponding tree.
To modify the tree to reflect the genealogy at a nearby location,
we simply remove those edges whose intervals do not overlap that location,
and add those new edges whose intervals do.
Incidentally, this property that edges naturally encode \emph{differences}
between nearby trees (e.g., as ``subtree prune and regraft'' moves)
allows for efficient algorithms to compute statistics of the genome sequence that take advantage
of the highly correlated nature of nearby trees~\citep{kelleher2016efficient}.

Given the topology defined by the nodes and edges, \emph{sites} and \emph{mutations}
encode the sequence information for each sample in an efficient way. Each site
is associated with a position on the genome and an ancestral state. For example,
in Figure~\ref{fig:example_tree_sequence} we have two sites, one at position
2 with ancestral state `A' and the other at position 7 with ancestral state `G'. If
no mutations occur, all samples inherit the ancestral state at a given site.
Each mutation occurs above a specific node at a given site,
and results in a specific derived state.
Thus, all samples below the mutation's node in the tree will inherit this state,
unless further mutations are encountered.
Three mutations are shown in Figure~\ref{fig:example_tree_sequence},
illustrated by red hexagons.
The first site, in the left-hand tree,
has a single mutation, which results in node $2$ inheriting the state `T'.
The second site, in the right hand tree, has two mutations:
one occurring over node $3$ changing the state to `C',
and a back mutation over node $1$ changing the state to `G'.

@@

\paragraph{Validity of a set of tables}
Given a set of node and edge tables as described above,
there are only two requirements that ensure the tables
describe a valid tree sequence.
% (There are essentially no such requirements on the site and mutation tables.)
These are:
\begin{enumerate}
    \item Offspring must be born after their parents (and hence, no loops).
    \item The set of intervals on which each individual is a child must be disjoint.
\end{enumerate}
A pair of node and edge tables that satisfy these two requirements
is guaranteed to uniquely describe at each point on the genome
a collection of directed, acyclic graphs -- in other words, a forest of trees.
For some applications it is necessary to check that at every point
there is only a \emph{single} tree.
Checking this is more difficult, but is implemented in \msprime{}'s API.
% \jda{TODO add function name}  PLR: not sure this is necessary.
For efficiency, \msprime{} makes several other sortedness requirements on the tables,
that are not necessarily satisfied by tables emitted by a forwards-time simulation.
\msprime{}'s API includes tools to rectify this by first sorting % (using \texttt{sort\_tables})
and then using the \texttt{simplify} algorithm described below, which works on sorted tables
and is guaranteed to produce a valid, \msprime{}-ready tree sequence.

@@

To record the genealogical history of a forwards in time simulation
we need to record two things for each new chromosome:
the birth time, and the endpoints and parental IDs of each distinctly inherited segment,
which are naturally stored as the \emph{nodes} and \emph{edges} of a tree sequence.
For concreteness, we write out in pseudocode how to run a neutral Wright--Fisher simulation
that records genealogical history in this way.
The simulation will run for $T$ generations,
and has $N$ haploid individuals, each carrying a single chromosome of length $L$.
For simplicity, we sample exactly one crossover per generation.

We use $\randomuniform(A)$ to denote an element of the set $A$ chosen uniformly at random.
% (and all such instances is independent).  % this is understood here
Given a node table $\Nt$, the function $\taddnode{\Nt, t}$ 
adds a new node to the table $\Nt$ with time $t$
and returns the ID of this new node. 
Similarly, the function $\taddedge{\Et, \ell, r, p, c}$ 
adds a new edge $(\ell, r, p, c)$ to the edge table $\Et$.
The function $\tsimplify{P, \Nt, \Et}$ (described in section \ref{ss:simplify})
simplifies the history stored in the tables $\Nt$ and $\Et$ 
to the minimal information required to represent the genealogies of the list of node IDs $P$;
after simplification the nodes appearing in $P$ are relabeled $(0, 1, \ldots, |P|)$.
A step-by-step explanation follows the pseudocode.

\begin{taocpalg}{W}{Forwards-time tree sequence}
{Simulates a randomly mating population of $N$ haploid individuals with
chromosome of length $L$ for $T$ generations, and returns the node
and edge tables ($\Nt$ and $\Et$) recording the simulated history.
The tables are simplified every $s$ generations, removing genealogical
information from $\Nt$ and $\Et$ unreachable from the current population $P$.
}

\algstep{W1.}{Initialisation.}{
Set $\Nt \leftarrow \mbox{NodeTable}()$, $\mathcal{E}
\leftarrow \mbox{EdgeTable}()$, $t \leftarrow T$, and $j \leftarrow 0$.
 For $0 \leq k < N$, set $P_k \leftarrow \taddnode{\Nt, T}$. 
}

\algstep{W2.}{Generation loop head: new node.}{Set $u \leftarrow \taddnode{\Nt, t}$ and $P'_j \leftarrow u$.
}

@@

\algstep{W6.}{Simplify.}{Call $\tsimplify{P', \Nt, \Et}$, and set $P'_k
\leftarrow k $ for $0 \leq k < N$. (Tables may need to be sorted.)}

@@

We begin in~\algref{W1} by creating new node and edge tables, and setting
our population $P$ (a vector of $N$ node IDs) to the initial population.
This initial population is a set of $N$ nodes with birth time $T$ generations
ago. We also initialise our generation clock $t$ and individual index $j$.
Steps~\algref{W2}--\algref{W4} replace the individual with node ID
$P_j$ by first creating a new node with birth time $t$ (and ID $u$).
In step~\algref{W3} we choose two indexes $a$ and $b$ uniformly, 
giving us parental IDs $P_a$ and $P_b$, 
and choose a chromosomal breakpoint $x$. 
We record the effects of this event by storing two new edges: one recording that the parent of node $u$
from $0$ to $x$ is $P_a$, and another recording that the parent of $u$
from $x$ to $L$ is $P_b$. Step \algref{W5} then iterates these steps
for each of the $N$ individuals for each generation. 
At the end of a generation, we then check
if we need to simplify (which is done every $s$ generations). 
If simplification is required, we do this in step \algref{W6} by calling the simplify function
on the node and edge tables with the current set of population IDs $P'$ as the samples. 
This updates the tables in-place to remove all redundant
information, and remaps the specified sample IDs to $0, \dots, N - 1$ in the updated tables. 
Hence, we set our current population IDs to
$0, \dots N - 1$ after simplify has completed.  
Step \algref{W7} loops these steps until the required number of generations have been simulated.
% then completes the algorithm by looping over generations; 
% we decrement our clock $t$, and terminate if $t = 0$.
% Otherwise, we update our current population and individual index and loop
% back to \algref{W2}.
